Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-27 threshold/draft.Rnw
Fix inequalities

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 \ge 2(1-t)$, then we can conclude from
just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.
In practice, many pairs of vectors may be pruned in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.

@@

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_{1} -
e_j^Tv_{1j}|^2 \ge 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
longest run of successive points in $P v_1$ within the interval
$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
419}$) from Section \ref{examples} in increasing order.  The lines in the plot
illustrate the interval associated with the correlation threshold $t=0.95$. The
quantity $\ell$ is obtained by rolling the interval across all the points in
the plot and counting the maximum number of points that lie within the
interval.  With respect to the ordering $P$, $\ell$ represents the biggest
index difference that paris of indices corresponding to correlated vectors
above the threshold can exhibit.

@@

Even a 1-d projected interval distance threshold may prune many possible
non-adjacent vector pairs with respect to the ordering defined by $P$ as
illustrated by the example in Figure \ref{fig1}.  However, this observation is
unlikely to rule out many adjacent pairs of vectors in most problems. For
example, the minimum distance between the points shown in the example in Figure
\ref{fig1} is less that $\sqrt{2(1-t)}/\sigma_1$, and no pruning at all occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors with respect to the ordering defined by $P$
below the correlation threshold. For example,
including only one more term finds that 49\% of adjacent vectors fall below the
a correlation threshold of $0.95$ in the example shown in Figure \ref{fig1} and
are pruned by identifying indices such that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
leaving only 15 candidate pairs of adjacent vectors that might possibly
meet the correlation threshold.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

>>>>

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 > 2(1-t)$, then we can conclude from
just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.
In practice, many pairs of vectors may be pruned in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.

@@

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_{1} -
e_j^Tv_{1j}|^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
longest run of successive points in $P v_1$ within the interval
$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
419}$) from Section \ref{examples} in increasing order.  The lines in the plot
illustrate the interval associated with the correlation threshold $t=0.95$. The
quantity $\ell$ is obtained by rolling the interval across all the points in
the plot and counting the maximum number of points that lie within the
interval.  With respect to the ordering $P$, $\ell$ represents the biggest
index difference that paris of indices corresponding to correlated vectors
above the threshold can exhibit.

@@

Even a 1-d projected interval distance threshold may prune many possible
non-adjacent vector pairs with respect to the ordering defined by $P$ as
illustrated by the example in Figure \ref{fig1}.  However, this observation is
unlikely to rule out many adjacent pairs of vectors in most problems.  For
instance, the minimum distance between the points shown in the example in
Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors
below the correlation threshold
with respect to the ordering defined by $P$.
For example,
including only one more term finds that 49\% of adjacent vectors fall below the
a correlation threshold of $0.95$ in the example shown in Figure \ref{fig1} and
are pruned by identifying indices such that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
leaving only 15 candidate pairs of adjacent vectors that might possibly
meet the correlation threshold.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.
